---
output: github_document
authors: Rose Birkner, Owen Senn
---

# Impact of Health Factors on Diabetes

## Rose Birkner, Owen Senn

#### Introduction

This project's goal is to explore and analyze the impact that certain health and risk factors have on diabetes. This topic is of interest because diabetes has been a growing problem in the United States for years and can usually be prevented by lifestyle choices. It is important to understand what steps individual's can take to mitigate and prevent this disease.

It is important to note that while the survey does not differentiate between type 1 and type 2 diabetes the vast majority of diabetes cases are type 2. According to the Centers for Disease Control and Prevention, "More than 38 million Americans have diabetes (about 1 in 10), and about 90% to 95% of them have type 2 diabetes." While it is possible to have a genetic predisposition to diabetes, type 2 is usually preventable through lifestyle choices.

#### Questions to be addressed

-   How do health factors such as blood pressure and BMI affect an individuals chances of getting diabetes? 
-   Does having diabetes put an individual at risk for having a stroke or heart disease/attack?
-   How do physical activity and diet (eating fruits and vegetables) affect the likelihood of getting diabetes? Is one significantly more influential than the other?
-   What is the relationship between mental health and diabetes? Is poor mental health associated with poor physical activity levels and/or diet which in turn affects the occurrence of diabetes?

#### Data Description

The data set for this project came from kaggle here: <https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/data?select=diabetes_012_health_indicators_BRFSS2015.csv>. That is a cleaned version of a data set which came from another kaggle project here: <https://www.kaggle.com/datasets/cdc/behavioral-risk-factor-surveillance-system?select=2015.csv>. This data comes from the Behavioral Risk Factor Surveillance System (BRFSS), an annual phone survey conducted by the CDC. Every year the data is collected from over 400,000 adults across America. The purpose of the survey is to collect information on possible health indicators of several chronic diseases as well as to record general health trends. During the survey, 330 variables are recorded. Among the information gathered is demographics, physical activity level, diet, chronic diseases, and more. Our focus for this project is looking at Diabetes and the related health factors that might indicate the occurrence of it.

Due to the large number of variables and observations, the original data set would have been impractical to use and unmanageable. Additionally, because the survey records such a wide variety of information it was unnecessary to look at variables that weren't related to the focus of our project, diabetes. For these reasons we used the more selective dataset on kaggle, which was much more manageable and included the variables necessary to look at diabetes and the health factors that might influence it. On the kaggle project there were three csv files. We chose to use the file with over 250,000 observations, a variable that recorded whether someone was diabetic, prediabetic, or had no diabetes, and 21 other variables. This particular dataset was from the BRFSS survey conducted in 2015.

#### Data Cleaning

The data set we are using was previously cleaned by Alex Teboul on Kaggle with the notebook found here: <https://www.kaggle.com/code/alexteboul/diabetes-health-indicators-dataset-notebook>. He compacted the data from the original 330 columns to the 22 we have now. This cleaning renamed columns for readability, removed instances with NA values, and cleaned up values to have 0 = no and 1 = yes for binary variables.

All of the categorical variables were of numeric type, so we changed them all to be factors. In addition, we wanted to combine the Fruits and Veggies column so that we could look at one variable that reflected diet. We called this column Fruits_and_Veggies and gave it 3 levels, 2 for an individual that said they eat both fruits and vegetables, 1 for an individual that eats either fruits or vegetables but not both, and 0 for someone that says they eat neither. The code for those modifications is below.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)

health_data <- read.csv("HealthFactor.csv")

#make categorical variables of type factor
health_data <- health_data %>% 
  mutate(
    Diabetes_012 = as.factor(Diabetes_012),
    HighBP = as.factor(HighBP),
    HighChol = as.factor(HighChol),
    CholCheck = as.factor(CholCheck),
    Smoker = as.factor(Smoker),
    Stroke = as.factor(Stroke),
    HeartDiseaseorAttack = as.factor(HeartDiseaseorAttack),
    PhysActivity = as.factor(PhysActivity),
    Fruits = as.factor(Fruits),
    Veggies = as.factor(Veggies),
    HvyAlcoholConsump = as.factor(HvyAlcoholConsump),
    AnyHealthcare = as.factor(AnyHealthcare),
    NoDocbcCost = as.factor(NoDocbcCost),
    GenHlth = as.factor(GenHlth),
    DiffWalk = as.factor(DiffWalk),
    Sex = as.factor(Sex),
    Age = as.factor(Age),
    Education = as.factor(Education),
    Income = as.factor(Income)
  )
  


#add fruits_and_veggies feature
health_data <- health_data %>% mutate(
  Fruits_and_Veggies = if_else(Fruits == 1 & Veggies == 1, 2,
                               if_else(Fruits == 1 | Veggies == 1, 1, 0)),
  Fruits_and_Veggies = as.factor(Fruits_and_Veggies)
) %>% 
  relocate(
    Diabetes_012:Veggies,
    Fruits_and_Veggies
  )
```

#### Variables

-   **Diabetes_012:** categorical variable where 0 is for no diabetes or only during pregnancy, 1 is for prediabetes, and 2 is for diabetes.

-   **HighBP:** categorical variable where 0 indicates no high blood pressure and 1 indicates that an individual has been told they have high blood pressure by a doctor, nurse, or other health professional.

-   **HighChol:** categorical variable where 0 indicates no high cholesterol and 1 indicates that an individual has been told by a doctor, nurse or other health professional that their blood cholesterol is high.

-   **CholCheck:** categorical variable where 0 indicates an individual has not had a cholesterol check in the last 5 years and 1 is that they have.

-   **BMI:** numerical variable of an individual's body mass index.

-   **Smoker:** categorical variable where 0 indicates an individual has not smoked at least 100 cigarettes in their life and 1 indicates they have.

-   **Stroke:** categorical variable where 0 indicates no stroke for an individual and 1 indicates they have been told they had a stroke.

-   **HeartDiseaseorAttack:** categorical variable where 0 indicates respondents that have never reported having coronary heart disease (CHD) or myocardial infarction (MI) and 1 indicates they have.

-   **PhysActivity:** categorical variable where 0 indicates no physical activity in the last 30 days besides an individual's regular job and 1 indicates physical activity or exercise during the past 30 days other than their regular job.

-   **Fruits:** categorical variable where 0 indicates no fruits eaten during the day and 1 indicates fruit consumption one or more times per day.

-   **Veggies:** categorical variable where 0 indicates no vegetables eaten during the day and 1 indicates vegetable consumption one or more times per day.

-   **Fruits_and_Veggies:** categorical variable where 0 indicates no fruits or vegetables eaten during the day, 1 indicates eating some fruit or some vegetables every day, and 2 indicates eating both fruits and vegetables every day.

-   **HvyAlcoholConsump:** categorical variable where 1 indicates a heavy drinker and 0 is not a heavy drinker. Heavy drinker is defined as adult men having more than 14 drinks per week and adult women having more than 7 drinks per week.

-   **AnyHealthcare:** categorical variable where 1 indicates having any kind of health care coverage and 0 is no healthcare coverage.

-   **NoDocbcCost:** categorical variable. 1 = yes and 0 = no in response to the question "Was there a time in the past 12 months when you needed to see a doctor but could not because of cost?"

-   **GenHlth:** categorical variable with 5 levels that indicates general health. 1 = excellent, 2 = very good, 3 = good, 4 = fair, 5 = poor.

-   **MentHlth:** numerical variable that indicates the number of days in the last 30 days that the participant's mental health was not good.

-   **PhysHlth:** numerical variable that indicates the number of days in the last 30 days that the participant's physical health was not good.

-   **DiffWalk:** categorical variable. 1 = yes and 0 = no in response to the question "Do you have serious difficulty walking or climbing stairs?"

-   **Sex:** categorical variable. 0 = female, 1 = male.

-   **Age:** categorical variable with 13 levels that indicate the age range of the participant. 1 = 18-24, 2 = 25-29, 3 = 30-34, 4 = 35-39, 5 = 40-44, 6 = 45-49, 7 = 50-54, 8 = 55-59, 9 = 60-64, 10 = 65-69, 11 = 70-74, 12 = 75-79, 13 = 80 or older.

-   **Education:** categorical variable with 6 levels that indicates education level. 1 = never attended or only kindergarten, 2 = grades 1-8 (Elementary), 3 = grades 9-11 (some high school), 4 = grade 12 or GED (high school graduate), 5 = college 1-3 years (some college or technical school), 6 = college 4 years or more (college graduate).

-   **Income:** categorical variable with 8 levels that indicate income level. 1 = less than \$10,000, 2 = less than \$15,000, 3 = less than \$20,000, 4 = less than \$25,000, 5 = less than \$35,000, 6 = less than \$50,000, 7 = less than \$75,000, 8 = \$75,000 or more

## Results

```{r}
library(ggplot2)
```

#### How do health factors such as blood pressure and BMI affect an individuals chances of getting diabetes? 

For this question, we wanted to observe and analyze how health factors such as blood pressure and BMI affect the chance of an individual having diabetes. This analysis is important to understand the benefits of living a healthy lifestyle and the consequences that may come if not. While there are some instances where blood pressure levels can be uncontrollable, most people can mitigate and control their blood pressure and BMI by living a healthy lifestyle. As a reminder, blood pressure is a categorical variable where 0 indicates no high blood pressure and 1 indicates that an individual has been told they have high blood pressure by a doctor, nurse, or other health professional. BMI is a numerical variable that represents a person's body mass index. This analysis will exam each variable on its own. 

First, we can graphically analyze how blood pressure relates to diabetes.

```{r}
health_data |> ggplot(aes(x = HighBP, fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Blood Pressure Evaluation", x = "Levels of blood pressure", y = "percentage", fill = "Diabetes status")
```

From this chart, we can see that individuals who have high blood pressure are way more likely to have diabetes than individuals who do not have high blood pressure. This difference is major and important to note and for people to understand. 

Next, we can examine the more controllable of the two variables BMI.

```{r}
health_data |>
  ggplot(aes(x = Diabetes_012, y = BMI, fill = Diabetes_012)) + geom_boxplot(outliers = FALSE) + scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")) +labs(title = "BMI Evaluation", x = "Diabetes levels", y = "BMI", fill = "Diabetes status")

geom_boxplot()
  
```

From this chart, we can see that individuals who have a higher BMI are at risk for diabetes. Due to this finding, it is important that people try to remain healthy to protect themselves from a higher BMI or high blood pressure, which can result in diabetes and many other chronic diseases. 

Due to BMI being a controllable variable, what can people do to improve it?
```{r, fig.show='hold', out.width="50%"}
health_data |>
  ggplot(aes(x = PhysActivity, y = BMI, fill = PhysActivity)) + geom_boxplot(outliers = FALSE) + scale_fill_manual(values = c("#66C2A5", "#8DA0CB"), labels = c("0" = "Not Active", "1" = "Active")) + labs(title = "BMI and Physical Activity Evaluation", x = "Physical Activity levels", y = "BMI", fill = "Activity status")

health_data |>
  ggplot(aes(x = Fruits, y = BMI, fill = Fruits)) + geom_boxplot(outliers = FALSE) + scale_fill_manual(values = c("#66C2A5", "#8DA0CB"), labels = c("0" = "No Fruits Eaten", "1" = "Fruits Eaten")) +labs(title = "BMI and Fruit Consumption Evaluation", x = "Fruit levels", y = "BMI", fill = "Fruit status")

health_data |>
  ggplot(aes(x = Veggies, y = BMI, fill = Veggies)) + geom_boxplot(outliers = FALSE) + scale_fill_manual(values = c("#66C2A5", "#8DA0CB"), labels = c("0" = "No Veggies Eaten", "1" = "Veggies Eaten")) +labs(title = "BMI and Veggie Consumption Evaluation", x = "Veggie levels", y = "BMI", fill = "Veggie status")


```

These charts show that your BMI will decrease by being more active and eating healthier. By doing these things, people will become healthier, which will lead them to be less prone to diabetes. By being healthier, blood pressure will also decrease, which also impacts diabetes. This is a specific and actionable plan that people can take to make sure they are protecting themselves against chronic diseases such as diabetes. 

#### Does having diabetes put an individual at risk for having a stroke or heart disease/attack?

While this project has examined how certain variables impact diabetes, it is also important to understand how having diabetes can impact your health. This question will explore whether diabetes puts a person at higher risk for a stroke or heart disease/attack. This is very important as these things can be life-threatening, which makes this issue of diabetes even more serious. The variable stroke is a categorical variable where 0 indicates no stroke for an individual and 1 indicates they have been told they had a stroke. The variable HeartDiseaseorAttack is also a categorical variable where 0 indicates respondents who have never reported having coronary heart disease (CHD) or myocardial infarction (MI) and 1 indicates they have. These variables will be analyzed on their own and also compared with age.

The variable stroke will first be analyzed.
```{r}
health_data |> ggplot(aes(x = Stroke, fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Stroke Evaluation", x = "Levels of Stroke", y = "percentage", fill = "Diabetes status")
```

As seen in this chart, the risk of getting a stroke while having diabetes is much higher. This is a very scary but also important insight, as it provides individuals' motivation to stay healthy and to be aware of the effects of diabetes. 

How does this vary by age?
```{r}
health_data |> ggplot(aes(x = Stroke, fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Stroke Evaluation by Age", x = "Levels of Stroke", y = "percentage", fill = "Diabetes status") + facet_wrap(~Age)
```

This idea that diabetes puts you at risk for a stroke remains consistent throughout each age group, along with some other alarming statistics. Individuals in their late teens and early twenties who would never have to worry about having a stroke because of their age now have to be aware that this is a possibility due to diabetes. 

The variable regarding heart disease will be analyzed next. 
```{r}
health_data |> ggplot(aes(x = HeartDiseaseorAttack, fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Heart Disease/Attack Evaluation", x = "Levels of Heart Disease/Attack", y = "percentage", fill = "Diabetes status")
```

As seen in this chart, the risk of having a heart attack or heart disease while having diabetes is much higher and similar to that of a stroke. Again, this is a very scary but also important insight.

How does this vary by age?
```{r}
health_data |> ggplot(aes(x = HeartDiseaseorAttack, fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Heart Disease/Attack Evaluation by Age", x = "Levels of Heart Disease/Attack", y = "percentage", fill = "Diabetes status") + facet_wrap(~Age)
```

Similar to the stroke graphs, the idea that diabetes has an impact on heart disease/attacks can be seen in every age group. Also, note that younger individuals are at risk, which is typically unheard of at this age.  

Diabetes is a very serious disease, and individuals need to be aware of the health risks it can bring in order to stay healthy. Being healthy is the most important thing to help prevent this disease and the risks that are associated with it. 

#### How do physical activity and diet (eating fruits and vegetables) affect the likelihood of getting diabetes? Is one significantly more influential than the other?

For this question we wanted to look at how diet and physical activity affect the rates of diabetes and whether prioritizing one over the other could lead to different chances of getting diabetes. In this survey the only features that pertain to diet are the variables Fruits and Veggies. Recall from the description of these variables that they are both categorical with 0 indicating no fruits or vegetables eaten during the day and 1 indicating fruit or vegetable consumption one or more times per day. As stated in the data cleaning section, we thought it would be interesting to combine these two variables so we could easily compare those who eat both fruits and vegetables at least once per day to those that only prioritize eating one as well as those who eat neither fruits nor vegetables. The details of how we defined this variable are in the data cleaning section. When diet is mentioned we are referring to the variable Fruits_and_Vegetables.

This first graphic is a look at basic fruit consumption and diabetes as well basic vegetable consumption and diabetes

```{r, fig.show='hold', out.width="50%"}
health_data %>% ggplot(aes(x = Fruits, fill = Diabetes_012)) +
  geom_bar() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Fruit and Diabetes status", x = "Level of fruit consumption", y = "count", fill = "Diabetes status")

       
health_data %>% ggplot(aes(x = Veggies, fill = Diabetes_012)) +
  geom_bar() +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"),labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Vegetables and Diabetes status", x = "Level of veggie consumption", y = "count", fill = "Diabetes status")
```

We can see from looking at those plots that visualizing the difference in proportions is difficult. After making this observation, we decided to go with a percent stacked bar chart to better visualize the difference in proportions. Those percent stacked bar charts are below.

```{r, fig.show='hold', out.width="50%"}
health_data %>% ggplot(aes(x = Fruits, fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Fruit and Diabetes status", x = "Level of fruit consumption", y = "percentage", fill = "Diabetes status")

       
health_data %>% ggplot(aes(x = Veggies, fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"),labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Vegetables and Diabetes status", x = "Level of veggie consumption", y = "percentage", fill = "Diabetes status")
```

This next plot is the combined fruits and vegetables feature compared with diabetes, which we also plotted as a percent stacked bar chart.

```{r}
health_data %>% ggplot(aes(x = Fruits_and_Veggies, fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" =  "Prediabetic", "2" = "Diabetic")) +
  labs(title = "Fruit & Veggies vs Diabetes status", x = "Level of fruit and veggie consumption", y = "percentage", fill = "Diabetes status")
```

With the percent stacked bar charts we can see much clearer that for both the fruit and vegetable plot the proportion of those with diabetes is lower for those that do incorporate fruits and vegetables into their diet compared to those that do not. Similarly, we see in the combined fruits and vegetables plot that the rate of diabetes is lower for those that consume both fruits and veggies on a daily basis than for those that include only fruits or only veggies on a daily basis. Therefore, a combination of fruits and vegetables in the diet is associated with a lower rate of diabetes.

Next we wanted to look at the relationship between physical activity and diabetes, which we can see in the next plot.

```{r}
health_data %>% ggplot(aes(PhysActivity, fill = Diabetes_012)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Physical activity and Diabetes status", x = "Physical activity level", y = "percentage", fill = "Diabetes status")
```

While there was a noticeable decline in the proportion of those with diabetes when looking at those that eat fruit and/or vegetables compared with those who don't, we can see from this plot that the difference in proportions is much more noticeable when looking at physical activity compared to the diet variables. This suggests that physical activity could potentially have a bigger impact on diabetes than diet does. The next graph looks at the relationship between diet, physical activity and diabetes together.

```{r}
health_data %>% mutate(
  Fruits_and_Veggies = factor(Fruits_and_Veggies, levels = c(0, 1, 2), labels = c("none", "fruits or veggies", "fruits and veggies"))
) %>% ggplot(aes(x = PhysActivity, fill = Diabetes_012)) +
  facet_wrap(~ Fruits_and_Veggies) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"),labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Diet, Physical Activity and Diabetes status", x = "Physical activity level", y = "Percentage", fill = "Diabetes status")
```

Here we can see that even for those with a poor diet (eating neither fruits nor vegetables) including physical activity into their lifestyle could significantly decrease their chances of developing diabetes. Not surprisingly, the category with the lowest proportion of diabetics is the category that both incorporates fruits and vegetables into their diet regularly and are physically active.

While it is true that those with a good diet who are physically active have the lowest rates of diabetes, it seems that the dominating factor here is physical activity. Diet definitely plays a role in this based on the first graphs we looked at, but physical activity seems to have a slightly more noticeable affect.


#### What is the relationship between mental health and diabetes? Is poor mental health associated with poor physical activity levels and/or diet which in turn affects the occurrence of diabetes?

To answer this question we first looked at the relationship between mental health and diabetes with a histogram and a box plot. When we first made these visualizations we saw that an overwhelming number of respondents said they had zero days of poor mental health in the past month. This made it difficult to read and interpret the rest of the data. We made the decision to exclude those who had zero days of poor mental health from these visualizations so that we were only looking at the set of people from the survey who had experienced some amount of poor mental health. For the remainder of exploration for this question we will only be looking at those with at least one day of poor mental health. Below are the first two plots.

```{r, fig.show='hold', out.width="50%", warning=FALSE, message=FALSE}
health_data %>% filter(
  MentHlth != 0
) %>% ggplot(aes(x = MentHlth, fill = Diabetes_012)) +
  geom_histogram(color = "black") +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Days of Poor Mental Health and Diabetes Level",x = "Num days poor mental health in last 30 days", y = "count", fill = "Diabetes status")


health_data %>% filter(
  MentHlth != 0
) %>% 
  ggplot(aes(x = Diabetes_012, y = MentHlth, fill = Diabetes_012)) +
  geom_boxplot(staplewidth = 1) +
  scale_fill_manual(values = c("#66C2A5", "#8DA0CB","#FC8D62"), labels = c("0" = "Not Diabetic", "1" = "Prediabetic", "2" = "Diabetic")
  ) +
  labs(title = "Days of Poor Mental Health and Diabetes Level", x = "Diabetes level", y = "Num days poor mental health in last 30 days", fill = "Diabetes status")
```

Some interesting observations can be made from these plots. In the histogram we can see that there is a spike of responses at about five or less days of poor mental health. after this spike we see the number of responses slowly declining as the number of days increases. At 30 days however, we see another spike of responses which includes the highest count of diabetes across all days in the graph. The boxplot shows something very similar. Those without diabetes have the lowest median for number of poor mental health days, those who are pre-diabetic tended to report a higher number of days and have a higher median, and finally, those who are diabetic tended to report the highest number of poor mental health days and have the highest median.

These plots suggest that there is a relationship between mental health and diabetes. However, they may not be directly related. Something else we wanted to explore in this question was whether or not mental health was related to some other factors that in turn would cause diabetes. In this next plot we look at the relationship between mental health and diet to see if diet could possibly be a link between mental health and diabetes.

```{r, fig.show='hold', out.width="50%", warning=FALSE, message=FALSE}
health_data %>% filter(
  MentHlth != 0
) %>% ggplot(aes(x = MentHlth, fill = Fruits_and_Veggies)) +
  geom_histogram(color = "black") +
  scale_fill_manual(values = c("#e78ac3", "#a6d854","#ffd92f"), labels = c("0" = "No Fruits or Veggies", "1" = "Some Fruits or Veggies", "2" = "Both Fruits and Veggies")
  ) +
  labs(title = "Days of Poor Mental Health and Diet", x = "Num days poor mental health in last 30 days", y = "count", fill = "Quality of Diet")


health_data %>% filter(
  MentHlth != 0
) %>% 
  ggplot(aes(x = Fruits_and_Veggies, y = MentHlth, fill = Fruits_and_Veggies)) +
  geom_boxplot(staplewidth = 1) +
  scale_fill_manual(values = c("#e78ac3", "#a6d854","#ffd92f"), labels = c("0" = "No Fruits or Veggies", "1" = "Some Fruits or Veggies", "2" = "Both Fruits and Veggies")
  ) +
  labs(title = "Days of Poor Mental Health and Diet", x = "Quality of Diet", y = "Num days poor mental health in last 30 days", fill = "Quality of Diet")
```

We can see from the plots that those with a poor diet of no fruits and vegetables tend to report a higher number of days of poor mental health. As we've already seen, diet plays a very important role in the likelihood of one developing diabetes. Therefore, this connection to a poor diet could factor in to the relationship between mental health and diabetes.

In the next plot we will look at the relationship between mental health and physical activity levels.

```{r, fig.show='hold', out.width="50%", warning=FALSE, message=FALSE}
health_data %>% filter(
  MentHlth != 0
) %>% ggplot(aes(x = MentHlth, fill = PhysActivity)) +
  geom_histogram(color = "black") +
  scale_fill_manual(values = c("#fb8072", "#80b1d3"), labels = c("0" = "No phys activity in last 30 days", "1" = "Phys activity in last 30 days")) +
  labs(title = "Days of Poor Mental Health and Physical Activity", x = "Num days poor mental health in last 30 days", y = "count", fill = "Level of physical activity")


health_data %>% filter(
  MentHlth != 0
) %>% ggplot(aes(x = PhysActivity, y = MentHlth, fill = PhysActivity)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#fb8072", "#80b1d3"), labels = c("0" = "No phys activity in last 30 days", "1" = "Phys activity in last 30 days"))
```

From this plot it is very clear that those who are not physically active tend to report far more days of poor mental health compared to those who are physically active. we know that physical activity influences the development of diabetes, and now we can see a clear relationship between mental health and physical activity level. Like we concluded from the last plots when considering diet, the connection between mental health and physical activity level could be another reason why poor mental health is associated with higher rates of diabetes.

#### Conclusion

This project's goal was to analyze different health and lifestyle factors that could potentially influence the development of diabetes. Health factors that we found associated with diabetes were high blood pressure and a higher BMI. Lifestyle factors that we found associated with diabetes were a poor diet (lack of fruits and vegetables) as well as a lack of physical activity. Not surprisingly, the combination of physical activity and a good diet showed a low rate of diabetes. It is important that people are aware of this as they are lifestyle choices that they can control. We also found that poor mental health was associated with diabetes, which could be due to the connection between mental health, diet, and physical activity. It is important that people are aware that diabetes is a serious condition with serious consequences. One alarming thing we found was that people with diabetes are much more likely to have a stroke or heart problems, even at a young age. However, if we are aware of the risks and the ways in which they can be avoided, we can take the steps necessary to prevent the development of diabetes.




